Supercritical fluid extraction of essential oil from chamomile seeds: experiments, modelling, and parameter estimation

### 2024-02-06 SFE_article_Parameter_Estimation - Chamomile/Sections/Model.tex
06.02

	{\footnotesize
		\begin{equation} \label{EQ:Velocity}
			\int \frac{\partial ({\color{black}\rho_f}({\color{black}T}(t,z),{\color{black}P}(t)) v {\color{black}A \phi}(z) )}{\partial {\color{black}z}} dz = 0 \rightarrow {\color{black}F}={\color{black}\rho_f}({\color{black}T}(t,z),{\color{black}P}(t)) v {\color{black}A\phi}(z)
		\end{equation}
	}

@@

			{\footnotesize
				\begin{align} \label{EQ:Enthalpy_equation}
					&\cfrac{\partial \left({\color{black}\rho_f}({\color{black}T}(t,z),{\color{black}P}(t)) {\color{black}h}(t,z) A_f\right)}{\partial t} - \cfrac{\partial \left({\color{black}P}(t) A_f\right)}{\partial t} \nonumber \\
					&+ \cfrac{\partial \left( {\color{black}\rho_f}({\color{black}T}(t,z),{\color{black}P}(t)) {\color{black}h}(t,z) A_f v \right)}{\partial {\color{black}z}} - \cfrac{\partial}{\partial {\color{black}z}} \left( k \cfrac{\partial {\color{black}T}(t,z)}{\partial {\color{black}z}} \right)
				\end{align}
			}

@@

		\subsubsection{Initial and boundary conditions} 
		It is assumed that the solvent is free of solute at the entrance of the extractor and that all the solid particles have the same initial solute content ${\color{black}c_{s0}}$. Every SFE system needs some time to reach the desired operating conditions but the solute diffuses from solid phase to the fluid phase already at the preparation stage. For instance, a pump introduce more fluid to the extractor to increase the pressure, which makes the fluid present already in that vessel to moves internally hence the solute in the fluid phase is non-uniformly distributed. Some conclusions can be drawn from the analysis of the initial part of each yield curve obtained from the laboratory (Figure \ref{fig: estimation_results}). It can be noticed that each curve at the beginning has a non-linear curvature. A quadratic function could approximate the initial part of each extraction curve. A function that, after integration, gives a quadratic-like result is a straight line. Based on that observation, the solute concentration in the fluid phase is assumed to be linearly distributed. The solute concentration is assumed to be zero at the outlet and reach the maximum at the beginning of the fixed bed. The details on the calculation are given in Appendix \ref{CH: IC_BC}. The linear distribution $H(z)$ can be defined if the total mass of solute $m_{total}$ and initial mass ratio between solid and fluid phases $\tau$ are known.
		Moreover, it is considered that the initial temperature of the extractor in every place is the same and described by ${\color{black}h_0}$. Therefore, the initial conditions employed in the simulation are:

		{\footnotesize
				\begin{equation*}
					{\color{black}c_f}(t = 0, z) = H(z) \quad {\color{black}c_s}(t = 0, z) = {\color{black}c_{s0}} \quad {\color{black}h}(t = 0, z) = {\color{black}h_0}
				\end{equation*} }

@@

			In a state-space sense, the state variables of the system are the local concentrations of solute in the fluid and solid phases (${\color{black}c_f}(t,z)$ and ${\color{black}c_s}(t,z)$, respectively), and the local enthalpy of the pseudo-homogeneous phase (${\color{black}h}(t,z)$). The controllable input variables are the mass flow-rate and temperature of the solvent in the feed and the pressure in the extractor. Additionally, the pressure change is augmented with the state-space and denoted as ${\color{black}P}(t)$.
			The system state-space is extended by assuming that extraction yield can be modelled as a function of a known initial mass of solute in the solid phase and it can be measured after the separator (${\color{black}Y}(t)$). The system is controllable by manipulating the flow-rate and temperature (enthalpy) of $CO_2$ in the feed, and the pressure in the extractor.

>>>>

	{\footnotesize
		\begin{equation}
			\int \frac{\partial ({\color{black}\rho_f}({\color{black}T}(t,z),{\color{black}P}(t)) v {\color{black}A \phi}(z) )}{\partial {\color{black}z}} dz = 0 \rightarrow {\color{black}F}={\color{black}\rho_f}({\color{black}T}(t,z),{\color{black}P}(t)) v {\color{black}A\phi}(z)
		\end{equation}
	}

@@

			{\footnotesize
				\begin{align} \label{EQ:Enthalpy_equation}
					\cfrac{\partial \left({\color{black}\rho_f}({\color{black}T}(t,z),{\color{black}P}(t)) {\color{black}h}(t,z) A_f\right)}{\partial t} &= - \cfrac{\partial \left( {\color{black}\rho_f}({\color{black}T}(t,z),{\color{black}P}(t)) {\color{black}h}(t,z) A_f v \right)}{\partial {\color{black}z}}  \nonumber \\
					& + \cfrac{\partial \left({\color{black}P}(t) A_f\right)}{\partial t} + \cfrac{\partial}{\partial {\color{black}z}} \left( k \cfrac{\partial {\color{black}T}(t,z)}{\partial {\color{black}z}} \right)
				\end{align}
			}

@@

		\subsubsection{Initial and boundary conditions} 
		It is assumed that the solvent is free of solute at the entrance of the extractor and that all the solid particles have the same initial solute content ${\color{black}c_{s0}}$. As the residence time is much shorter than the sampling time, the initial state estimate for the concentration of the solute in the fluid phase would be not reliable. Considering that it is assumed that the $c_f0=0$. Moreover, it is considered that the initial temperature of the extractor in every place is the same and described by ${\color{black}h_0}$. Therefore, the initial conditions employed in the simulation are:

		{\footnotesize
				\begin{align*}
					&{\color{black}c_f}(t = 0, z) = 0 & {\color{black}c_s}(t = 0, z) = {\color{black}c_{s0}} & &{\color{black}h}(t = 0, z) = {\color{black}h_0} \\
					&\frac{\partial c_f(t,z=L)}{\partial x} = 0 & {\color{black}c_s}(t = 0, z) = {\color{black}c_{s0}} & &\frac{\partial h(t,z=L)}{\partial x} = 0
				\end{align*} }

@@

			In a state-space sense, the state variables of the system are the local concentrations of solute in the fluid and solid phases (${\color{black}c_f}(t,z)$ and ${\color{black}c_s}(t,z)$, respectively), and the local enthalpy of the pseudo-homogeneous phase (${\color{black}h}(t,z)$). The controllable input variables are the mass flow-rate and temperature of the solvent in the feed and the pressure in the extractor. Additionally, the pressure change is augumented with the state-space and denoted as ${\color{black}P}(t)$.
			The system state-space is extended by assuming that extraction yield can be modelled as a function of a known initial mass of solute in the solid phase and it can be measured after the separator (${\color{black}Y}(t)$). The system is controllable by manipulating the flow-rate and temperature (enthalpy) of $CO_2$ in the feed, and the pressure in the extractor.

### 2024-02-06 SFE_article_Parameter_Estimation - Chamomile/Sections/Parameter_estimation.tex
06.02

	{\footnotesize
		\begin{equation} \label{EQ: Objective_function}
			\ln L = -\cfrac{n}{2} \left( \ln \sqrt{2 \pi} + \ln {\color{black} \sigma}^2 \right)
			- \cfrac{ \sum_{i=1}^{n} \left[  \cfrac{d {Y}(t_i)}{dt} - \cfrac{d{\color{black}y}({\color{black} \theta}, t_i)}{dt} \right]^2 }{2 {\color{black} \sigma}^2}
		\end{equation}
	}

@@

	\begin{itemize}
		\item Partition coefficient:\qquad\quad\qquad$k_m$
		\item Internal diffusion coefficient: \quad$D_i^R$
		\item Axial diffusion coefficient: \qquad$D_e^M$
		\item Decay coefficient:\qquad\qquad\qquad$\Upsilon$
		\item Standard deviation:\qquad\qquad\quad~$\sigma^2$
		\item Total mass of solute: \qquad\quad\quad$m_{total}$
		\item Initial mass ratio: \qquad\qquad\qquad$\tau$
	\end{itemize}

@@

 	\begin{table}[h]
		\adjustbox{max width=\columnwidth}{%
			\begin{tabular}{ lccccccc }
				\hline 
				Parameter		&$k_m$[-] 	& $D_i^R$[$m^2/s$] 	& $D_e^M$[$m^2/s$]	& $\Upsilon$ [m$^3$/kg] & $m_{total}$[g]	& $\tau$[-] 	& $\sigma$[-] \\  \hline
				Lower bound		&0	  		& 0 	  			& 0 				&	0					& 80 		 		& 0 	   		& 0 \\ 
				Upper bound		&$+\infty$	& $+\infty$ 		& $+\infty$			&	100					& 150 				& 1 			& $+\infty$ \\ 
				Initial guess	&1e5 		& 3 	  			& 1 				&	0					& 80 		 		& 0.65 	   		& 0.1 \\  \hline
		\end{tabular} }
		\caption{Constraints and initial guess}
		\label{tab:Constraints}
	\end{table}

>>>>

	{\footnotesize
		\begin{equation} \label{EQ: Objective_function}
			\ln L = -\cfrac{n}{2}  \ln \sqrt{2 \pi \sigma^2} 
			- \cfrac{ \sum_{i=1}^{n} \left[  \cfrac{d {Y}(t_i)}{dt} - \cfrac{d{\color{black}y}({\color{black} \theta}, t_i)}{dt} \right]^2 }{2 {\color{black} \sigma}^2}
		\end{equation}
	}

@@

	\begin{itemize}
		\item Partition coefficient:\qquad\quad\qquad$k_m$
		\item Internal diffusion coefficient: \quad$D_i^R$
		%\item Axial diffusion coefficient: \qquad$D_e^M$
		\item Decay coefficient:\qquad\qquad\qquad$\Upsilon$
		\item Standard deviation:\qquad\qquad\quad~$\sigma^2$
	\end{itemize}

@@

 	\begin{table}[h]
		\adjustbox{max width=\columnwidth}{%
			\begin{tabular}{ lccccccc }
				\hline 
				Parameter		&$k_m$[-] 	& $D_i^R\times10^{-13}$[$m^2/s$] 	& $D_e^M$[$m^2/s$]	& $\Upsilon$ [m$^3$/kg] & $\sigma$[-] \\  \hline
				Lower bound		&0	  		& 0 	  			& 0 				& 0 	   		& 0 \\ 
				Upper bound		&$+\infty$	& $+\infty$ 		& $+\infty$			& $+\infty$ 	& $+\infty$\\ 
				Initial guesses	&0.1-10		& $0.1-10$ 			& 0 		   		& 0.1-10		& 0.01 \\  \hline
		\end{tabular} }
		\caption{Constraints and initial guess}
		\label{tab:Constraints}
	\end{table}

### 2024-02-06 SFE_article_Parameter_Estimation - Chamomile/Sections/introduction_imp.tex
06.02

	This study aims to develop a process model for extracting natural substances from solid materials and liquids using supercritical fluids, specifically supercritical $CO_2$. To achieve this, the model parameters are estimated based on thermodynamic relations and a set of experiments conducted at various conditions. The maximum likelihood estimator is used to solve the parameter estimation problem, and the obtained parameters are subjected to regression to derive correlations. These correlations enable the process model to be generalized across a range of temperatures (40 to $50~^\circ C$) and pressures (200 to 300 bar).

	The study is structured as follows: Chapter \ref{CH: Thermodynamic} provides a general discussion on supercritical fluids to familiarize the reader with their properties. Chapter \ref{CH:Governing_equations_chapter} introduces the general balance equations. The balance equations are combined with the extraction kinetic equation to develop the process model in Chapter \ref{CH: Extraction_model}. The maximum likelihood technique is presented in Chapter \ref{CH: Parameter_estimation} and is then combined with the process model. Chapter \ref{CH: Experiments} describes the experimental work and the data collected from experiments are used to estimate missing parameters related to the extraction kinetic. Finally, the parameter estimations and simulation results are discussed in Chapters \ref{CH: Results} and \ref{CH: Conclusion}

>>>>

	This study aims to develop a process model for extracting natural substances from solid materials and liquids using supercritical fluids, specifically supercritical $CO_2$. To achieve this, the solvent properties are estimated based on thermodynamic relations and the extraction kinetic parameters estimated based on a set of experiments conducted at various conditions. The maximum likelihood estimator is used to solve the parameter estimation problem, and the obtained parameters are subjected to regression to derive correlations. These correlations enable the process model to be generalized across a range of temperatures (40 to $50~^\circ C$) and pressures (200 to 300 bar).

	The study is structured as follows: Chapter \ref{CH: Thermodynamic} provides a general discussion on supercritical fluids to familiarize the reader with their properties. Chapter \ref{CH:Governing_equations_chapter} introduces the general balance equations. The balance equations are combined with the extraction kinetic equation to develop the process model in Chapter \ref{CH: Extraction_model}. The maximum likelihood technique is presented in Chapter \ref{CH: Parameter_estimation} and is then combined with the process model. Chapter \ref{CH: Experiments} describes the experimental work and the data collected from experiments, which are used to estimate missing parameters related to the extraction kinetic. Finally, the parameter estimations and simulation results are discussed in Chapters \ref{CH: Results} and \ref{CH: Conclusion}

### 2024-02-06 SFE_article_Parameter_Estimation - Chamomile/Supercritical_fluid_extraction_of_essential_oil_from_chamomile.tex
06.02

% ===================================================
% METADATA
% ===================================================
\title[mode=title]{Supercritical fluid extraction of essential oil from caraway seeds: experiments, modelling, and parameter estimation}                      
\shorttitle{Supercritical fluid extraction of essential oil from caraway seeds: experiments, modelling, and parameter estimation}

@@

% ===================================================
% ABSTRACT
% ===================================================
\begin{abstract}
This study investigated the supercritical extraction process of caraway oil from caraway seeds. A distributed-parameter model describes the fluid-solid extraction process. The concept of quasi-one-dimensional flow is applied to reduce the number of spatial dimensions. The flow is assumed to be uniform across any cross-section, although the area available for the fluid phase can vary along the extractor. The physical properties of the solvent are estimated from the Peng-Robinson equation of state. Key model parameters included partition factor internal and axial diffusion coefficients. These parameters and the initial state were determined using maximum likelihood estimation from experimental data, assuming normal error. A set of laboratory experiments was performed in a partially filled extractor with a fixed bed operated under multiple constant operating conditions: $40~^\circ C$ / $200$ bar, $50~^\circ C$ / $200$ bar, $40~^\circ C$ / $300$ bar, and $50~^\circ C$ / $300$ bar.

@@

This study presented basic aspects regarding the laboratory work for obtaining caraway extracts using supercritical carbon dioxide at different operating conditions. A first-principle process model was developed based on governing equations. As a result of the literature review, the extraction kinetic model was selected and combined with the general mass balance equations. The values of unknown parameters in the process model were obtained by parameter estimation. The yield data from the experiments were compared against the yield generated by the model. The IPOPT solver was used to solve the maximum likelihood estimation. The process model was found to have good agreement with experimental yield curves. Later, the parameters found for each experiment were combined to obtain more general relationships. The correlations for diffusion and decay coefficients are presented as a function of fluid density or Reynolds number. It should be noted that obtained correlations have been prepared with a limited amount of data. These correlations can be used to study qualitatively the impact of different operating conditions on the extraction yield by different methods, such as sensitivity analysis. Moreover, the generalized model could be useful in finding the optimal operating conditions with respect to the process economy.

@@

\subsection{Initial and boundary conditions} \label{CH: IC_BC}
\subfile{Sections/IC_BC}

@@

\subsection{Solid density measurement} \label{CH: Solid_Density_Measurment}

Figure \ref{fig:density_cal} shows results of density measurement performed with pycnometer.

\begin{figure}[!h]
	\centering 
	\includegraphics[trim=2cm 6cm 4cm 0cm, clip,width=\columnwidth]{Sections/ultraReportT5.pdf}
	\caption{The result of solid density measurement}
	\label{fig:density_cal}
\end{figure}

{\footnotesize
	\begin{equation*}
		\rho_s^{ave} = \frac{1.2585+1.2582+1.2561+1.2546+1.2555}{5} = 1.25658 [g/cc]
	\end{equation*}
}

\subsection{Porosity calculations} \label{CH: Porosity}
\subfile{Sections/Porosity}

>>>>

% ===================================================
% METADATA
% ===================================================
\title[mode=title]{Supercritical fluid extraction of essential oil from chamomile seeds: experiments, modelling, and parameter estimation}                      
\shorttitle{Supercritical fluid extraction of essential oil from chamomile seeds: experiments, modelling, and parameter estimation}

@@

% ===================================================
% ABSTRACT
% ===================================================
\begin{abstract}
This study investigated the supercritical extraction process of chamomile oil from chamomile seeds. A distributed-parameter model describes the fluid-solid extraction process. The concept of quasi-one-dimensional flow is applied to reduce the number of spatial dimensions. The flow is assumed to be uniform across any cross-section, although the area available for the fluid phase can vary along the extractor. The physical properties of the solvent are estimated from the Peng-Robinson equation of state. Key model parameters included partition factor internal and axial diffusion coefficients were determined using maximum likelihood estimation from experimental data, assuming normal error. A set of laboratory experiments was performed in under multiple constant operating conditions: $30 - 40~^\circ C$, $100 - 200$ bar, and $1.33-6.66 \times 10^{-5}$ kg/s.

@@

This study presented basic aspects regarding the laboratory work for obtaining chamomile extracts using supercritical carbon dioxide at different operating conditions. A first-principle process model was developed based on governing equations. As a result of the literature review, the extraction kinetic model was selected and combined with the general mass balance equations. The values of unknown parameters in the process model were obtained by parameter estimation. The yield data from the experiments were compared against the yield generated by the model. The IPOPT solver was used to solve the maximum likelihood estimation. The process model was found to have good agreement with experimental yield curves. Later, the parameters found for each experiment were combined to obtain more general relationships. The correlations for diffusion and decay coefficients are presented as a function of fluid density or Reynolds number. It should be noted that obtained correlations have been prepared with a limited amount of data. These correlations can be used to study qualitatively the impact of different operating conditions on the extraction yield by different methods, such as sensitivity analysis. Moreover, the generalized model could be useful in finding the optimal operating conditions with respect to the process economy.

@@

%\subsection{Initial and boundary conditions} \label{CH: IC_BC}
%\subfile{Sections/IC_BC}

@@

%\subsection{Solid density measurement} \label{CH: Solid_Density_Measurment}

%Figure \ref{fig:density_cal} shows results of density measurement performed with pycnometer.

%\begin{figure}[!h]
%	\centering 
%	\includegraphics[trim=2cm 6cm 4cm 0cm, clip,width=\columnwidth]{Sections/ultraReportT5.pdf}
%	\caption{The result of solid density measurement}
%	\label{fig:density_cal}
%\end{figure}

%{\footnotesize
%	\begin{equation*}
%		\rho_s^{ave} = \frac{1.2585+1.2582+1.2561+1.2546+1.2555}{5} = 1.25658 [g/cc]
%	\end{equation*}
%}

%\subsection{Porosity calculations} \label{CH: Porosity}
%\subfile{Sections/Porosity}
